---
title: "PSTAT 122 - HW5"
author: "Emily Lu"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, echo=FALSE, message=FALSE}
library(knitr)
library(dplyr)
library(kableExtra)
library(tidyverse)
library(agricolae)
options(knitr.table.format = "latex")
```

__13.1__ An experiment was performed to investigate the capability of a measurement system. Ten parts were randomly selected, and two randomly selected operators measured each part three times. The tests were made in random order, and the data are shown below: 

```{r}
part <- factor(rep(1:10,times=6))
operator <- factor(rep(1:2,each=30))
measurement <- c(50, 52, 53, 49, 48, 52, 51, 52, 50, 47,
                 49, 52, 50, 51, 49, 50, 51 ,50, 51, 46,
                 50, 51, 50, 50, 48, 50, 51, 49, 50, 49,
                 50, 51, 54, 48, 48, 52, 51, 53, 51, 46,
                 48, 51, 52, 50, 49, 50, 50, 48, 48, 47,
                 51, 51, 51, 51, 48, 50, 50, 50, 49, 48)
head(cbind(part,operator,measurement))
```

(a) Analyze the data from this experiment. 

    Sol. $H_0: \; \sigma^2_{\tau \beta} = 0$  vs. $H_1$: $\sigma^2_{\tau \beta} > 0$

    ```{r}
meas.aov <- aov(measurement ~ part*operator)
summary(meas.aov)
```

    According to the ANOVA table above, the p-value (0.927) is greater than $\alpha = 0.05$ which means that there is insufficient evidence to suggest that the interaction between the parts and operators significantly affects the measurement. 
    Since the interaction was not significant, we will now test the individual factors. Our hypotheses will be $H_0:\; \sigma^2_\tau = 0$ vs $H_1: \; \sigma^2_\tau >0$. 
    
    ```{r}
a <- 10; b <- 2; n <- 3
F_part <- 11.002/0.602
pf(F_part, (a-1), (a-1)*(b-1), lower.tail = F)
```

    Since the p-value is very small (< $\alpha = 0.05$), we would reject $H_0$. There is sufficient evidence to suggest that the parts significantly affect measurement.
    
    ```{r}
    F_operator <- 0.417/0.602
    pf(F_operator,(a-1), (a-1)*(b-1), lower.tail = F)
    ```

    Since the p-value is $0.7034 > \alpha = 0.05$, fail to reject $H_0$. There is insufficient evidence to suggest that the operators significantly affect measurement.
    
(b) Estimate the variance components using the ANOVA method. 

    $\hat{\sigma}^2 = MSE = 1.500$

    $\hat{\sigma}^2_{\tau\beta} = \frac{0.602 - 1.500}{3} = -0.2993333$

    $\hat{\sigma}^2_{\tau} = \frac{11.002 - 0.602}{2*3} = 1.733333$

    $\hat{\sigma}^2_{\beta} = \frac{0.417 - 0.602}{10*3} = -0.006166667$

__13.5__ Reanalyze the measurement systems experiment in Problem 13.1, assuming that operators are a fixed factor. Estimate the appropriate model components using the ANOVA method.

Since operators is now a fixed factor and parts is still a random factor, this is now a mixed model. The interaction, however, is still a random effect. The ANOVA table is the same, but F-statistics may change.

$H_0: \beta_j = 0 \forall j$ vs. $H_1: \beta_j \neq 0$ for some $j$ 

$F_0 = \frac{MSB}{MSAB} = \frac{0.417}{0.602} = 0.692691$

```{r}
pf(0.692691, b-1, (a-1)*(b-1), lower.tail = F)
```

Since p-value = $0.4267823 > \alpha = 0.05$, we fail to reject $H_0$. 

$H_0: \sigma^2_\tau = 0$ vs. $H_1: \sigma^2_\tau > 0$ 

$F_0 = \frac{MSA}{MSE} = \frac{11.002}{1.500} = 7.334667$

```{r}
pf(7.334667, a-1, a*b*(n-1), lower.tail = F)
```

Since p-value = $0.0001103954 < \alpha = 0.05$, we reject $H_0$. 

$\hat{\sigma}^2_\tau = \frac{MSA-MSE}{bn} = \frac{11.002 - 1.500}{2*3} = 1.583667$

$\hat{\sigma}^2_{\tau \beta} = \frac{MSAB - MSE}{n} = \frac{0.602 - 1.500}{3} = -0.2993333$

$\hat{\sigma}^2 = MSE = 1.500$
