---
title: "PSTAT 174 & 274 - Homework 2"
author: "Emily Lu"
date: "10/24/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(astsa)
```

#### 2.2.a For the mortality data examined in Example 2.2: Add another component to the regression in (2.21) that accounts for the particulate count four weeks prior; that is, add $P_{t-4}$ to the regression in (2.21). State your conclusion.

Let $M_t = \beta_0 + \beta_1 t + \beta_2(T_t - T_\circ) + \beta_3(T_t - T_\circ)^2 + \beta_4P_t + w_t$ (2.21) be the reduced model and 
$M_t = \beta_0 + \beta_1 t + \beta_2(T_t - T_\circ) + \beta_3(T_t - T_\circ)^2 + \beta_4P_t + \beta_5P_{t-4} + w_t$ (2.22) be the full model that accounts for the particulate count four weeks prior. We test the reduced model (2.21) against the full model (2.22) by comparing their coefficient of determination. For the reduced model (2.21), we have the summary table, AIC, BIC, and AICs values below: 

```{r}
#  Regression
temp  <- tempr-mean(tempr)  # center temperature    
temp2 <- temp^2             # square it  
trend <- time(cmort)        # time
n <- length(cmort)

fit <- lm(cmort~ trend + temp + temp2 + part, na.action = NULL)
            
summary(fit) # regression results of 2.21
anova(fit)
AIC(fit)/n - log(2*pi)                                # AIC 
BIC(fit)/n - log(2*pi)                                # BIC 
(AICc <- log(sum(resid(fit)^2)/n) + (n+5)/(n-5-2)) # AICc
```
For the full model (2.2), we have the summary table, AIC, BIC, and AICs values below:  
```{r}
fit1 <- lm(cmort[5:n] ~ trend[5:n] + temp[5:n] + temp2[5:n] + part[5:n]
           + part[1:(n-4)], na.action = NULL)

summary(fit1)     # regression results of 2.21 with component that accounts for the particulate count four weeks prior
anova(fit1)

AIC(fit1)/n - log(2*pi)                                # AIC 
BIC(fit1)/n - log(2*pi)                                # BIC 
(AICc <- log(sum(resid(fit1)^2)/n) + (n+5)/(n-5-2)) # AICc
```
Comparing the results of both models, we see that the full model is better than the reduced model since all of its independent variables account for some 60.8% of the variability and it has the best value for AIC and BIC. 

#### 3.1 For an MA(1), $x_t = w_t + \theta w_{t-1}$, show that $|\rho_x(1)| \leq 1/2$ for any number $\theta$. For which values of $\theta$ does $\rho_x(1)$ attain its maximum and minimum? 

Solution: From example 3.5 in SS, we know that for MA(1), $\rho_x(1) = \frac{\theta}{1+\theta^2}$. To find the maximum and minimum, we would first take the derivative of $\rho_x(1)$, so $\frac{d\rho_x(1)}{d\theta} = \frac{1}{1+\theta^2} - \frac{2\theta^2}{(1+\theta^2)^2} = \frac{1-\theta^2}{(1+\theta^2)^2}$ then set it equal to 0. $\frac{d\rho_x(1)}{d\theta} = \frac{1-\theta^2}{(1+\theta^2)^2} = 0 \Longrightarrow \theta = \pm 1.$ At $\theta = 1$, $\rho_x(1) = \frac{1}{2}$ which is the maximum and for $\theta = -1$, $\rho_x(-1) = \frac{-1}{2}$. 

```{r, fig.align = 'center'}
par(cex.lab = 0.5, cex.axis = 0.5, mar=c(5,6,6,6))
curve(x / (1 + x^2), from = -1/2, to = 1/2, xlab = expression(theta), ylab = expression(over(theta, ( 1 + theta)^2)))
grid()
abline(v=0,h=0,lty=2)
```

#### Individual Problem: ARMA(p, q)

#### Consider an ARMA process whose AR polynomial is $\phi(z) = 1 - z^2/4$ and whose MA polynomial is $\theta(z) = 1 + z/3$. Complete the following: 

* Factor the AR and MA polynomials 

_Sol._ The AR polynomial can be factored as 
\begin{eqnarray} \phi(z) &=& 1 - z^2/4 \\ &=& 1^2 - (\frac{z}{2})^2 \\ &=& (1-\frac{z}{2})(1+\frac{z}{2}) \\ &=& \frac{1}{4}(2-z)(2+z). \end{eqnarray} The MA polynomial can be factored as \begin{eqnarray} \theta(z) &=& 1 + z/3 \\ &=& \frac{1}{3}(3 + z). \end{eqnarray}

* Determine where or not $\phi(z)$ and $\theta(z)$ share common factors. 

From the factoring of $\phi(z)$ and $\theta(z)$ above, we could see that $\phi(z)$ and $\theta(z)$ does not share common factors. $\phi(z)$'s factors are -2 and 2, while $\theta(z)$'s factor is -3. 

* For the ARMA(p,q) process corresponding to $\phi(z)$ and $\theta(z)$, what are $p$ and $q$?

The ARMA(p,q) process corresponding to $\phi(z)$ and $\theta(z)$ can be written as \begin{eqnarray} \phi(B)x_t = \theta(B)w_t &\Longleftrightarrow& (1 - \frac{1}{4}B^2)x_t = (1 + \frac{1}{3}B)w_t \\ &\Longleftrightarrow& x_t - \frac{1}{4}x_{t-2} = w_t + \frac{1}{3}w_{t-1} \\ &\Longleftrightarrow& x_t = \frac{1}{4}x_{t-2} + w_t + \frac{1}{3}w_{t-1}, \end{eqnarray} which is ARMA(2, 1) so we have that p = 2 and q = 1.  

* Is the ARMA(p,q) process corresponding to $\phi(z)$ and $\theta(z)$ invertible? Justifiy your answer. 

According to Property 3.2 in SS, an ARMA(p,q) process is invertible if and only if $\theta(z) \neq 0$ for $|z| \leq 1$ or equivalently, $\theta(z) = 0$ only when $|z| > 1$. Thus, to see if the ARMA(p,q) process corresponding to $\phi(z)$ and $\theta(z)$ is invertible, first set $\theta(z)$ equal to 0:  $\theta(z) = 1 + z/3 = 0  \Longleftrightarrow z = -3$. We could see that $\theta(z)$ equals 0 only when $z=-3$ which satisfies $\theta(z) = 0$ only when $|z| > 1$. Therefore, the ARMA(p,q) process corresponding to $\phi(z)$ and $\theta(z)$ is invertible.   

* Is the ARMA(p,q) process corresponding to $\phi(z)$ and $\theta(z)$ causal? Justifiy your answer. 

According to Property 3.1 in SS, an ARMA(p,q) process is causal if and only if $\phi(z) \neq 0$ for $|z| \leq 1$ or equivalently, $\phi(z) = 0$ only when $|z| > 1$. Thus, to see if the ARMA(p,q) process corresponding to $\phi(z)$ and $\theta(z)$ is causal, we set  $\phi(z) = 0$ and solve for $z$.  $\phi(z) = 1 - z^2/4 = 0 \Longleftrightarrow z = 2$. Since $\phi(z) = 0$ only when $|z| > 1$, this ARMA(p,q) process is causal. 