---
title: "PSTAT 122 - HW4"
author: "Emily Lu"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, echo=FALSE, message=FALSE}
library(knitr)
library(dplyr)
library(kableExtra)
library(tidyverse)
library(agricolae)
options(knitr.table.format = "latex")
```

__5.12__ An experiment is conducted to study the influence of operating temperature and three types of faceplate glass in the light output of an oscilloscope tube. The following data are collected:
```{r 1, message=FALSE, echo = FALSE, fig.align='center'}
type1 <- matrix(c(580, 1090, 1392, 568, 1087, 1389, 570, 1085, 1386), 3, byrow = TRUE, dimnames = list(c("", "", ""), c("100", "125", "150")))
type2 <- matrix(c(550, 1070, 1328, 530, 1035, 1312, 579, 1000, 1299), 3, byrow = TRUE, dimnames = list(c("", "", ""), c("100", "125", "150")))

df <- rbind(type1, type2)
kable(df) %>%
  kable_styling(position = "center") %>%
  add_header_above(c("Glass Type", "Temperature" = 3)) %>%
  pack_rows("1", 1, 3) %>%
  pack_rows("2", 4, 3)
```

(a) Use $\alpha = 0.05$ in the analysis. Is there a significant interaction effect? Does glass type or temperature affect the response? What conclusions can you draw?

    Sol. $H_0: (\tau \beta)_{ij} = 0$ for all $i, j$ vs. $H_1$: at least one $(\tau \beta)_{ij} \neq 0$
    
    ```{r}
light <- c(580, 1090, 1392, 568, 1087, 1380, 570, 1085, 1386,
        550, 1070, 1328, 530, 1035, 1312, 579, 1000, 1299)
glass <- as.factor(c(rep(1, 9), rep(2, 9)))
temperature <- as.factor(rep(c(rep(100, 1), rep(125, 1), rep(150, 1)), 6))
df <- cbind(glass, temperature, light)

resp.aov <- aov(light ~ glass*temperature)
summary(resp.aov)
```

    According to the ANOVA table above, there is not a significant interaction between the glass type and temperature since the p-value (0.084350) is greater than $\alpha = 0.05$. The main effects of the glass type and temperature, however, are significant since both of their p-values are less than $\alpha = 0.05$. 

    ```{r}
interaction.plot(temperature, glass, light)
```

    Since the lines in the interaction plot are almost parallel, this also suggests that there is not a significant interaction between the glass types and temperatures. Additionally, we could see from the interaction plot that glass type 1 on average outputs more light than glass type 2 outputs. 